Synthesizing Symmetric Lenses

### 2019-01-29 papers/SymmetricLenses/main.tex
Checkpoint and done for today

To obtain SREs from the plain regular expressions that we expect users to
supply, we use a heuristic that attempts to assign probability annotations that
give each string in the language equal probablity. Our algorithm could also take
as input a stochastic regular expression rather than a plain one,
which would allow
the user to specify a data-specific probability distribution, either manually
calculated or inferred from a large dataset. We found our heuristic
works well in practice, so we have not explored these options.

Now that we have a way to calculate likelihood, we need a way to search for
likely lenses. Previous work on
synthesis~\cite{augustsson-2004,osera+:pldi15,feser-pldi-2015,frankle+:popl16}
has shown that we can use types to dramatically restrict the search space and
improve the effectiveness of synthesis. In our setting, types are pairs of
stochastic regular expressions where each regular expression is semantically
equal to infinitely many others. We follow existing work on bidirectional
program synthesis, and split our algorithm into two communicating
synthesizers~\cite{maina+:quotient-synthesis}. The first synthesizer, \Expand,
navigates the space of semantically equivalent regular expressions by applying
rewrite rules that preserve both the semantics and the probability distributions
of those stochastic regular expressions. Our algorithm ranks pairs of stochastic
regular expressions by the number of rewrite rule applications required to
obtain the pair from the stochastic regular expressions given as input. It
passes the pairs off to the second synthesizer, \GreedySynth, in rank order with
the smallest first.

At a high level, \GreedySynth looks for lenses between a given pair of
stochastic regular expressions $\Regex$ and $\RegexAlt$ that have the highest
likelihood by performing a type-directed search. In more detail, \GreedySynth
first normalizes the stochastic regular expressions provided by \Expand into
\emph{stochastic DNF regular expressions}. These stochastic DNF regular
expressions are, intuitively, stochastic regular expressions with disjunctions
distributed over concatenations, where associativity information has been
removed. Then, \GreedySynth uses the syntax of these n-ary DNF regular
expressions to find highly normalized lenses, in a form we call \emph{simple
  symmetric n-ary DNF lenses}. These lenses include neither a composition
operator nor a type equivalence rule. These restrictions mean that each simple
symmetric n-ary DNF lens is typed by a single pair of stochastic n-ary DNF
regular expressions and \GreedySynth's search space for a given pair of regular
expressions is finite. The variations that \GreedySynth considers include
choosing between \IdentityLens{} and \Disconnect lenses and deciding how to
match multiple occurrences of the same kind of data, \EG, the two occurrences of
\lstinline{name} in the management and HR data formats (we want to map first
names to first names and last names to last names). \GreedySynth
yields a simple symmetric lens by converting the n-ary
forms into the binary forms provided in the surface language.

As an example of the partnership between \Expand and \GreedySynth,
consider the lens shown in Figure~\ref{fig:example_lens}. 
Inferring this lens requires \Expand to use a star-semiring axiom to
rewrite the regular expression \lstinline{("\n" . emp_salary)$^*$} within
\lstinline{emp_salaries} into 
\lstinline{"" | (("\n" . emp_salary). ("\n" . emp_salary)$^*$)} 
so that \GreedySynth can find the appropriate lens.

>>>>

To obtain SREs from the plain regular expressions that users write, we use a
heuristic that attempts to assign probability annotations that give each
string in the language equal probablity. Our synthesis algorithm could
equally well take as input a\bcp{pair of} stochastic regular expression rather than a
plain one, permitting users to specify a data-specific probability
distribution (either manually calculated or inferred from some dataset of
interest). However, we have found that our heuristic works well in practice,
so we have not explored these options.

Now that we have a way to calculate likelihood, we need a way to search for
likely lenses. Previous work~\cite{augustsson-2004, osera+:pldi15,
  feser-pldi-2015, frankle+:popl16} has shown that we can use types to
dramatically restrict the search space and improve the effectiveness of
synthesis. In our setting, types are pairs of stochastic regular
expressions\bcp{the rest of the sentence feels out of place (and it's
  grammatically confusing, suggesting that there might be an RE that is NOT
  semantically equivalent to infinitely many others)}
where each regular expression is semantically equivalent to infinitely many
others. Following existing work on bidirectional program
synthesis~\cite{maina+:quotient-synthesis}, we 
split our algorithm into two communicating
synthesizers. The first synthesizer\bcp{why
call this algorithm a synthesizer?},
\Expand, navigates the space of semantically equivalent regular expressions
by applying rewrite rules that preserve both the semantics and the
probability distributions of those stochastic regular expressions. This
algorithm ranks pairs of stochastic regular expressions by the number of
rewrite rule applications required to obtain each pair from the one given as
input. It passes the pairs off to the second 
synthesizer, \GreedySynth, in rank order with the smallest first.

Abstractly, \GreedySynth looks for highest-likelihood lenses between a given pair of
stochastic regular expressions $\Regex$ and $\RegexAlt$ by performing a
type-directed search. In more detail, \GreedySynth 
first normalizes the stochastic regular expressions provided by \Expand into
\emph{stochastic DNF regular expressions}---a constrained representation of
stochastic regular expressions with disjunctions  
distributed over concatenations and with associativity information\bcp{what
  is ``associativity information''??} 
removed.  \GreedySynth then uses the syntax of these n-ary DNF regular
expressions to find highly normalized\bcp{is ``highly normalized'' different
from ``normalized''?} lenses, in a form we call \emph{simple
  symmetric n-ary DNF lenses}. These lenses include neither a composition
operator nor a type equivalence rule. These restrictions mean that each simple
symmetric n-ary DNF lens is typed by a single pair of stochastic n-ary DNF
regular expressions, hence \GreedySynth's search space for a given pair of regular
expressions is finite. The variations that \GreedySynth considers include
choosing between \IdentityLens{} and \Disconnect lenses and deciding how to
match multiple occurrences of the same kind of data, \EG, the two occurrences of
\lstinline{name} in the management and HR data formats (we want to map first
names to first names and last names to last names\bcp{the relationship of
  the parenthesized comment to what comes before is not clear}). Finally, \GreedySynth
yields a simple symmetric lens by converting the n-ary
forms into the binary forms provided in the surface language.

%%%%%%%%%%%%%%% BCP stopped here

As an example of the partnership between \Expand and \GreedySynth,
consider the lens shown in Figure~\ref{fig:example_lens}. 
Inferring this lens requires \Expand to use a star-semiring axiom\bcp{are we
assuming readers know what that is?} to
rewrite the regular expression \lstinline{("\n" . emp_salary)$^*$} within
\lstinline{emp_salaries} into 
\lstinline{"" | (("\n" . emp_salary). ("\n" . emp_salary)$^*$)} 
so that \GreedySynth can find the appropriate lens.
